---
title: "energy consumption forecasting"
author: "Mohammed_Nihal"
date: "3/29/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### loading librarys

```{r message=FALSE}
library(RMySQL)
library(dplyr)
library(lubridate)
library(ggplot2)
library(ggpubr)
library(ggpmisc)
library(plotly)
library(tidyr)
library(ggfortify)
library(forecast)
```



### Connecting to the database 

```{r}
con <- dbConnect(MySQL(), user='deepAnalytics', password='Sqltask1234!', dbname='dataanalytics2018', host='data-analytics-2018.cbrosir2cswx.us-east-1.rds.amazonaws.com')
```


### List the tables contained in the database 

```{r}
dbListTables(con)
```
### Import the table from database

```{r warning=FALSE}
yr_2006<-dbGetQuery(con,'select * from yr_2006')
yr_2007<-dbGetQuery(con,'select * from yr_2007')
yr_2008<-dbGetQuery(con,'select * from yr_2008')
yr_2009<-dbGetQuery(con,'select * from yr_2009')
yr_2010<-dbGetQuery(con,'select * from yr_2010')
```


### Combine tables into one dataframe using dplyr

```{r}
energy_consumption <- bind_rows(yr_2006,yr_2007,yr_2008,yr_2009,yr_2010)
```

### Combine Date and Time attribute values in a new attribute column

```{r}
energy_consumption <-cbind(energy_consumption,paste(energy_consumption$Date,energy_consumption$Time), stringsAsFactors=FALSE)
```

### Give the new attribute in the 11th column a header name 

```{r}
colnames(energy_consumption)[11] <-"DateTime"
```

### Move the DateTime attribute within the dataset

```{r}
energy_consumption <- energy_consumption[,c(ncol(energy_consumption), 1:(ncol(energy_consumption)-1))]
```

### Create a dataset filtering unwanted variables

```{r}
power_cal=energy_consumption %>% 
  select(DateTime,Global_active_power,Global_reactive_power,Sub_metering_1,Sub_metering_2,Sub_metering_3,Date) %>% 
  mutate(remaing_power=Global_active_power*(1000/60)-Sub_metering_1-Sub_metering_2-Sub_metering_3) %>% 
  mutate(Global_active_power=Global_active_power*1000/60) %>% 
  mutate(Global_reactive_power=Global_reactive_power*1000/60)
```

### Convert DateTime from POSIXlt to POSIXct

```{r}
power_cal$DateTime <- as.POSIXct(power_cal$DateTime,format="%Y-%m-%d %H:%M:%S",tz="Europe/Paris")
```




### Create "year",month,day,week,etc attribute with lubridate

```{r}
power_cal$year <- year(power_cal$DateTime)
power_cal$month <- month(power_cal$DateTime)
power_cal$quarter<-quarter(power_cal$DateTime)
power_cal$day<-day(power_cal$DateTime)
power_cal$week<-week(power_cal$DateTime)
power_cal$hour<-hour(power_cal$DateTime)
power_cal$minute<-minute(power_cal$DateTime)
power_cal$weekday<-weekdays(power_cal$DateTime)


```

### Plot the graph of active power during the whole time period

```{r fig.height = 10, fig.width = 5}
power_cal %>% filter(year!='2006') %>% ggplot(aes(DateTime,Global_active_power))+geom_line()+facet_wrap(year~.,scales = 'free_x',ncol = 1)
```









### Plotiing submeter 1 reading during a day

```{r}
c<-power_cal %>% filter(year == 2008 & month == 1 & day == 9)
plot_ly(c,x = ~c$hour, y=~c$Sub_metering_1, type = 'scatter', mode = 'lines')
```







### Plot sub-meter 1, 2 and 3 with title, legend and labels - All observations


```{r}
c <- filter(power_cal, year == 2008 & month == 1 & day == 9 & (minute == 0 | minute == 10 | minute == 20 | minute == 30 | minute == 40 | minute == 50))
plot_ly(c,x = ~c$hour, y=~c$Sub_metering_1,name='Kitchen', type = 'scatter', mode = 'lines')%>% 
 add_trace(y=~c$Sub_metering_2,name='laundary',mode='lines') %>% 
 add_trace(y=~c$Sub_metering_3,name='Heater and Ac',mode='lines') %>% 
  layout(title = "Power Consumption January 9th, 2008",  xaxis = list(title = "Time"),yaxis=list(title = "Power (watt-hours)"))

```




### Pie chart of overall submeter reading during 12 to 17 hours

```{r}
power_cal %>% filter(year==2008,month==1,12<=hour,hour<=17) %>% select(Sub_metering_1,Sub_metering_2,Sub_metering_3)%>% 
  rename(kitchen=Sub_metering_1,laundry=Sub_metering_2,heater_and_AC=Sub_metering_3) %>% 
  gather("Stat", "Value") %>% 
  plot_ly(labels = ~Stat, values = ~Value, type = 'pie') %>% 
 layout(title = 'One month overall submeter reading during 12 to 17 hrs',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))
```


### Pie chart of sub meter usage in a day 

```{r}
power_cal %>% filter(year!=2006) %>% select(Sub_metering_1,Sub_metering_2,Sub_metering_3)%>% 
  rename(kitchen=Sub_metering_1,laundry=Sub_metering_2,heater_and_AC=Sub_metering_3) %>% 
  gather("Stat", "Value") %>% 
  plot_ly(labels = ~Stat, values = ~Value, type = 'pie') %>% 
 layout(title = 'Overall submeter reading',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))
```


### Pie chart of sub meter usage in whole year 

```{r}
power_cal %>% filter(year==2009) %>% select(Sub_metering_1,Sub_metering_2,Sub_metering_3) %>%  rename(kitchen=Sub_metering_1,laundry=Sub_metering_2,heater_and_AC=Sub_metering_3)%>% 
  gather("Stat", "Value") %>% 
  plot_ly(labels = ~Stat, values = ~Value, type = 'pie') %>% 
 layout(title = 'Overall submeter reading',
         xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
         yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))
```

### Bar diagram for the active power usage in 2007

```{r}
power_cal %>% filter(year==2007) %>% group_by(month) %>% summarise(Global_active_power=sum(Global_active_power)) %>% 
ggplot(aes(x=month,y=Global_active_power))+geom_col()
```












### Each submeter reading in each year

```{r message=FALSE}
power_cal %>%  group_by(year) %>% summarise(kitchen=sum(Sub_metering_1),laundry=sum(Sub_metering_2),
                                                                    heater_and_AC=sum(Sub_metering_3),remaing_power=sum(remaing_power))%>% 
  gather("Stat", "Value", -year) %>% 
  ggplot(aes(x = year, y = Value, fill = Stat)) +
  geom_col(position = "dodge")+ggtitle("Sub meter reading of each year")


```

### power usage in each quarter of year

```{r}
power_cal %>% filter(year!=2006) %>% group_by(quarter) %>% summarise(kitchen=mean(Sub_metering_1),laundry=mean(Sub_metering_2),
                                                                    heater_and_AC=mean(Sub_metering_3),remaing_power=mean(remaing_power))%>% gather("Stat", "Value", -quarter) %>% 
  ggplot(aes(x = quarter, y = Value, fill = Stat)) +
  geom_col(position = "dodge")+ggtitle("power usage in each quarter")


```






### power usage in each month of year

```{r}
power_cal %>% filter(year!=2006) %>% group_by(month) %>% summarise(kitchen=mean(Sub_metering_1),laundry=mean(Sub_metering_2),
                                                              heater_and_AC=mean(Sub_metering_3))%>% gather("Stat", "Value", -month) %>% 
  ggplot(aes(x = month, y = Value, fill = Stat)) +
  geom_col()+ggtitle("power usage in each month of year")

```







### Power usage in each day of a month

```{r}

c<-power_cal %>% filter(year==2008,month==1) 
plot_ly(c,x = ~c$day, y=~c$Sub_metering_1,name='Kitchen', type = 'scatter', mode = 'lines')%>% 
 add_trace(y=~c$Sub_metering_2,name='laundary',mode='lines') %>% 
 add_trace(y=~c$Sub_metering_3,name='Heater and Ac',mode='lines') %>% 
  layout(title = "Power Consumption January, 2008",  xaxis = list(title = "Time"),yaxis=list(title = "Power (watt-hours)"))
```



### power usahe during a week

```{r}

c<-power_cal %>% filter(year==2008,month==1,week==2) 
plot_ly(c,x = ~c$weekday, y=~c$Sub_metering_1,name='Kitchen', type = 'bar')%>% 
 add_trace(y=~c$Sub_metering_2,name='laundary') %>% 
 add_trace(y=~c$Sub_metering_3,name='Heater and Ac') %>% 
  layout(title = "Power Consumption in each day",  xaxis = list(title = "day"),yaxis=list(title = "Power (watt-hours)"))
```





### power usage in tuesday

```{r}

c<-power_cal %>% filter(year==2008,month==1,week==2,weekday=='Tuesday') 
plot_ly(c,x = ~c$hour, y=~c$Sub_metering_1,name='Kitchen', type = 'bar')%>% 
 add_trace(y=~c$Sub_metering_2,name='laundary') %>% 
 add_trace(y=~c$Sub_metering_3,name='Heater and Ac') %>% 
  layout(title = "Power Consumption January, 2008",  xaxis = list(title = "Time"),yaxis=list(title = "Power (watt-hours)"))
```




### power usage in saturday

```{r}

c<-power_cal %>% filter(year==2008,month==1,day==14) 
plot_ly(c,x = ~c$hour, y=~c$Sub_metering_1,name='Kitchen', type = 'bar')%>% 
 add_trace(y=~c$Sub_metering_2,name='laundary') %>% 
 add_trace(y=~c$Sub_metering_3,name='Heater and Ac') %>% 
  layout(title = "Power Consumption January, 2008",  xaxis = list(title = "Time"),yaxis=list(title = "Power (watt-hours)"),barmode='overlay')
```




















### time series for the sub meter 1 reading 

```{r}
a=power_cal %>% filter(year!=2006) %>% 
 group_by(week,month,year) %>% summarise(Sub_metering_1=mean(Sub_metering_1))
tswsb1=ts(a$Sub_metering_1,frequency = 52,start = c(2007,1),end = c(2010,11))
autoplot(tswsb1,xlab = "Time", ylab = "Watt Hours", main = " sub_meter_1 ")
```



### linear regression forcasting for sub meter 1

```{r}
ftswsb1 <- tslm(tswsb1~ trend + season)
forecastfitSM3c <- forecast(ftswsb1, h=20, level=c(80,90))
plot(forecastfitSM3c, ylab= "Watt-Hours", xlab="Time")
```


### Decompose components for sub meter 1

```{r}

componentsb1<-decompose(tswsb1)
plot(componentsb1)
```




### Ets method forcasting for submeter 1

```{r}

tswsb1 %>% forecast( h=32,level=c(80,95), fan=FALSE) %>% autoplot()

  
```


### time series for the sub meter 2 reading 

```{r}
a=power_cal %>% filter(year!=2006) %>% 
 group_by(week,month,year) %>% summarise(Sub_metering_2=mean(Sub_metering_2))
tswsb2=ts(a$Sub_metering_2,frequency = 52,start = c(2007,1),end = c(2010,11))
autoplot(tswsb2, ts.colour= 'red', xlab = "Time", ylab = "Watt Hours", main = " sub_meter_2 ")
```


### Ets forecast for sub meter 2 

```{r}
tswsb2 %>% forecast( h=32,level=c(80,95), fan=FALSE) %>% autoplot()

```

### Decompose components of submeter2 
```{r}

componentsb1<-decompose(tswsb2)
plot(componentsb1)
```


### time series for the sub meter 3 reading 

```{r}
a=power_cal %>% filter(year!=2006) %>% 
 group_by(week,month,year) %>% summarise(Sub_metering_3=sum(Sub_metering_3))
tswsb3=ts(a$Sub_metering_3,frequency = 52,start = c(2007,1),end = c(2010,11))
autoplot(tswsb3, xlab = "Time", ylab = "Watt Hours", main = " sub_meter_3 ")
```




### Decompose components of submeter 3

```{r}
componentsb3<-decompose(tswsb3)
plot(componentsb3)
```


### Ets forecast for sub meter 3 

```{r}
tswsb3 %>% forecast( h=32,level=c(80,95), fan=FALSE) %>% autoplot()
```


### time series for the remaining meter reading 

```{r}
a=power_cal %>% filter(year!=2006) %>% 
 group_by(week,month,year) %>% summarise(remaing_power=sum(remaing_power))
tswr=ts(a$remaing_power,frequency = 52,start = c(2007,1),end = c(2010,11))
autoplot(tswr, xlab = "Time", ylab = "Watt Hours", main = " Remaining power")
```



### Decompose components of remaining power usage

```{r}

componentsb1<-decompose(tswr)
plot(componentsb1)
```
### create timeseries of overall power usage in months 
```{r}
a=power_cal %>% filter(year!=2006) %>% 
 group_by(month,year) %>% summarise(Global_active_power=mean(Global_active_power))
tsmap=ts(a$Global_active_power,frequency = 12,start = c(2007,1),end = c(2010,11))
autoplot(tsmap,  xlab = "Time", ylab = "Watt Hours", main = "active power ")
```


### linear regression forcast of overall power usage 

```{r}
ftsm <- tslm(tsmap~ trend + season)
forecastfitSM3c <- forecast(ftsm, h=20, level=c(80,90))
plot(forecastfitSM3c, ylim = c(0, 30), ylab= "Watt-Hours", xlab="Time") 
```


### Holt trend method for overall active power usage

```{r}
fc <- holt(tsmap, h=12)
fc2 <- holt(tsmap, damped=TRUE, phi = 0.9, h=12)
autoplot(tsmap) +
  autolayer(fc2, series="Damped Holt's method", PI=FALSE) +
  ggtitle("Forecasts from Holt's method") + xlab("year") +
  ylab("overall power usage ") +
  guides(colour=guide_legend(title="Forecast"))
```

### Holt winter method for overall active power usage

```{r}
fit1 <- hw(tsmap,damped = TRUE,seasonal="additive")
fit2 <- hw(tsmap,damped = TRUE, seasonal="multiplicative")
autoplot(tsmap) +
  autolayer(fit1, series="HW additive forecasts", PI=FALSE) +
  autolayer(fit2, series="HW multiplicative forecasts",
    PI=FALSE) +
  xlab("Year") +
  ylab("power usage") +
  ggtitle("overall power usage") +
  guides(colour=guide_legend(title="Forecast"))
```

### seasonal plot of active power

```{r}
ggseasonplot(tsmap, year.labels=TRUE, year.labels.left=TRUE) +
  ylab("power usage") +
  ggtitle("Seasonal plot: active power")
```



